---
title: "Comparing Shape Descriptors: CSS, Wavelets, and Elliptic Fourier"
author: "Víctor Manuel Tuset & Antoni Lombarte"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Shape Descriptors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  out.width = "100%",
  fig.retina = 2,
  dpi = 300
)
```

```{r setup}
library(aforoR)
library(ggplot2)
library(dplyr)
```

## About this tutorial

In otolith shape analysis, choosing the right mathematical descriptor is crucial for capturing the biological information relevant to your study. The `aforoR` package provides three main methods: **Elliptic Fourier Descriptors (EFD)**, **Wavelet Transform**, and **Curvature Scale Space (CSS)**. This vignette explains their mathematical foundations, advantages, and typical use cases.

## Overview Comparison

| Descriptor | Domain | Best For... | Invariance |
|:---|:---|:---|:---|
| **Elliptic Fourier (EFD)** | Frequency (global) | General shape classification, reconstruction | Translation, Rotation, Scale |
| **Wavelets** | Time-Frequency (local) | High-frequency details, stock separation | Translation, Scale (if normalized) |
| **Curvature Scale Space (CSS)** | Multi-scale Geometry | Feature detection, contour irregularities | Rotation, Translation, Uniform Scale |

```{r load_real_data}
# Load a sample otolith image from the package
img_path <- system.file("extdata", "otolith.jpg", package = "aforoR")
binary_img <- preprocess_image(img_path)
contour <- extract_contour(binary_img)

# Professional visualization theme
aforo_theme <- function() {
  theme_minimal() +
    theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 8, color = "black"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      axis.line = element_line(color = "black", linewidth = 0.3),
      axis.ticks = element_line(color = "black", linewidth = 0.3)
    )
}
```

---

## 1. Elliptic Fourier Descriptors (EFDs)

EFDs represent a closed contour as a sum of ellipses (harmonics). Each harmonic $k$ is defined by four coefficients ($a_k, b_k, c_k, d_k$).

- **Mathematics**: Decomposes the periodic functions $x(t)$ and $y(t)$ into Fourier series.
- **Interpretation**: 
  - **Low Harmonics (1-5)**: Capture the "global" shape, symmetry, and general elongation of the otolith (e.g., separating oval vs. rectangular shapes).
  - **High Harmonics (>15)**: Represent fine-grained margin details, such as small serrations or irregularities that can be specific to certain stocks or environments.
- **Typical Use**: Evolutionary studies, morphospace analysis, and species identification (Kuhl & Giardina, 1982).

```{r efd_example}
# Example: EFD is typically calculated using the Momocs package
# The aforoR package focuses on wavelets and CSS for otolith analysis
cat("EFD captures shape through harmonics:\n")
cat("- Harmonic 1-2: Elliptical outline\n")
cat("- Harmonic 3-5: Triangular/rectangular features\n")
cat("- Harmonic 6-10: Moderate detail\n")
cat("- Harmonic >15: Fine serrations\n")
```

---

## 2. Wavelet Transform

Wavelets analysis performs a multi-resolution decomposition of the contour signal (radius or step-length).

- **Mathematics**: Uses a mother wavelet $\psi(t)$ to analyze the signal at different scales $j$. Scale $j$ corresponds to a frequency band.
- **Interpretation**:
  - **High Scales (7-9)**: Capture the main body and coarsest features of the otolith contour.
  - **Mid Scales (4-6)**: Correspond to morphological features like lobes, indentations, and the general complexity of the protrusions.
  - **Low Scales (1-3)**: Represent the highest frequencies, often capturing very fine margin texture or high-frequency noise.
- **Typical Use**: Stock separation where differences occur in specific regions of the otolith margin (Parisi-Baradad et al., 2005).

```{r wavelets_example}
dists <- calculate_distances(contour)
wavelets <- calculate_wavelets_analysis(dists, n_scales = 9)

# Plot Scale 5
scale_5_df <- data.frame(Index = seq_len(ncol(wavelets$polar)), Value = wavelets$polar[5, ])
ggplot(scale_5_df, aes(x = Index, y = Value)) +
  geom_line(color = "coral2") +
  labs(title = "Wavelet Scale 5 (Polar distances)", x = "Point Index", y = "Coefficient") +
  aforo_theme()
```

---

## 3. Curvature Scale Space (CSS)

CSS tracks the evolution of inflection points on the contour as it's smoothed by a Gaussian kernel $G(t, \sigma)$.

- **Mathematics**: Inflection points are found where the second derivative of the smoothed curve is zero: $\kappa(t, \sigma) = 0$.
- **Interpretation**:
  - **CSS Image Peaks**: The highest peaks in the $(\sigma, t)$ space represent the most stable, "deepest" concavities or convexities that survive significant smoothing.
  - **Wide Distribution of Points at Low $\sigma$**: Indicates a contour with many fine-grained irregularities and high-frequency "wiggliness".
- **Typical Use**: Quantifying protrusions/lobes (Mokhtarian & Mackworth, 1986).

```{r css_example}
css_res <- calculate_css(contour)
# The plot method for css object already uses a professional style if configured
plot(css_res)
```

## Summary: Which one should I use?

1.  **Use EFD** for broad morphological studies where global shape is the primary driver of variation.
2.  **Use Wavelets** for detecting subtle population-level differences in margin complexity.
3.  **Use CSS** for identifying and quantifying specific geometric landmarks like lobes or indentations.

## References

- Kuhl, F.P. and Giardina, C.R. (1982). Elliptic Fourier features of a closed contour. *Computer Graphics and Image Processing*.
- Mokhtarian, F. and Mackworth, A.K. (1986). Scale-based description and recognition of planar curves and two-dimensional shapes. *IEEE Transactions on Pattern Analysis and Machine Intelligence*.
- Parisi-Baradad, V. et al. (2005). Otolith shape contour analysis using mathematical descriptors. *Computers in Biology and Medicine*.
